Since its inception in 1970, EPA's mission has been to pursue a
cleaner, healthier environment for the American people. The Agency was
assigned the daunting task of repairing the damage already done to the
natural environment and establishing new criteria to guide Americans
in making a cleaner environment a reality. Since 1970, the EPA has
worked with federal, state, tribal, and local partners to advance its
mission to protect human health and the environment. In order to carry
out its mission, EPA employs and collaborates with some of the
nation's best scientific minds. EPA prides itself in applying sound
science and state of the art techniques and methods to develop and
test innovations that will protect both human health and the
environment.

Under existing laws and recent Homeland Security Presidential
Directives, EPA has been called upon to play a vital role in helping
to secure the nation against foreign and domestic enemies.  The
National Homeland Security Research Center (NHSRC) was formed in 2002
to conduct research in support of EPA's role in homeland
security. NHSRC research efforts focus on five areas: water
infrastructure protection, threat and consequence assessment,
decontamination and consequence management, response capability
enhancement, and homeland security technology testing and
evaluation. EPA is the lead federal agency for drinking water and
wastewater systems and the NHSRC is working to reduce system
vulnerabilities, prevent and prepare for terrorist attacks, minimize
public health impacts and infrastructure damage, and enhance recovery
efforts. 

This Users Manual for the TEVA-SPOT Toolkit software package is published and
made available by EPA's Office of Research and Development to
assist the user community and to link researchers with their clients. 
\vspace*{\baselineskip}
\vspace*{\baselineskip}

\begin{tabular}{ll}
 & Jonathan Herrmann, Director \\
\\
\\
& National Homeland Security Research Center \\
& Office of Research and Development \\
& U. S. Environmental Protection Agency
\end{tabular}
