This section describes how to simulate contamination incidents and compute contamination impacts, which are the first steps needed to setup and solve a sensor placement problem with SPOT. These two steps can be viewed as preprocessing or data preparation for sensor placement optimization. Thus, these steps can be performed prior to optimization, which is generally a more interactive, iterative process.

The following sections illustrate the capabilities of SPOT with the example in the C:$\backslash$spot$\backslash$examples$\backslash$simple directory.

\section{Simulating Contamination Incidents}\label{impacts_impact1}

To simulate contamination incidents, the \doxyref{tevasim}{p.}{tevasimExecutable} command is utilized, which uses EPANET to perform an ensemble of contaminant transport simulations defined by a \doxyref{TSG File}{p.}{formats_tsgFile}. An ensemble of contamination scenarios for EPANET Example Net3 is defined in the file C:$\backslash$spot$\backslash$examples$\backslash$simple$\backslash$Net3.tsg. Contamination incidents are simulated for all network junctions, one for each hour of the day, and each contamination incident models an injection that continues for 24 hours. The {\ttfamily tevasim} command is run with the following command line:
\lstinputlisting[firstline=9,lastline=9]{examples/simple2.sh}% anchor=tevasim
This command generates several files: a binary ERD database that contains the contamination transport data (the database is stored in Net3.erd, Net3.index.erd, Net3-1.hyd.erd, and Net3-1.qual.erd), and Net3.out, which provides a textual summary of the EPANET simulations and is the same as the report file ($\ast$.rpt) from EPANET.

Adding \doxyref{DVF File}{p.}{formats_dvfFile} for flushing control allows the user to simulate a flushing response to an event. {\ttfamily tevasim} behaves slightly differently in this case, adding in the flushing rate to demands at the specified nodes, and closing the specified pipes. The number of pipes and/or nodes can be zero, if you don't want to use that part of the response policy.

To simulate contamination incidents using multi-species reactions, {\ttfamily tevasim} uses the EPANET multi-species extension (MSX).  EPANET-MSX is an extension to EPANET that enables complex reaction schemes between multiple chemical and biological species in both the bulk flow and at the pipe wall.  For a review of EPANET-MSX, see Shang, Uber and Rossman ~\cite{ShaUbeRos11}. Multi-species contamination incidents require a msx file and species declaration.  The following {\ttfamily tevasim} command simulates multi-species contamination incidents:
\lstinputlisting[firstline=9,lastline=9]{examples/simple7.sh}% anchor=tevasim 

\section{Using tso2Impact}\label{impacts_impact2}

After running \doxyref{tevasim}{p.}{tevasimExecutable} command, the output database, {\ttfamily Net3.erd}, can be used to compute one or more IMPACT files. An IMPACT file summarizes the consequence of each contamination incident in a manner that facilitates optimization. The \doxyref{tso2Impact}{p.}{tso2ImpactExecutable} command generates these files with the following command line: 
\lstinputlisting[firstline=16,lastline=16]{examples/simple2.sh}% anchor=tso2Impact
This command generates IMPACT files for each of the five objectives specified: mass consumed (mc), volume consumed (vc), time to detection (td), number of failed detections (nfd) and extent of contamination (ec). For each IMPACT file (e.g. {\ttfamily Net3\_\-mc.impact} ), a corresponding {\ttfamily ID} file is generated to map the sensor placement ids back to the network junction labels (e.g. {\ttfamily Net3\_\-mc.impact.id} ).

The impact measures computed by {\ttfamily tso2Impact} represent the
amount of impact that would occur up until the point where a contamination
incident is detected. This computation assumes that sensors work perfectly
(i.e., there are no false positive or false negative errors). However, the sensor behavior can be generalized in two ways. First, a detection threshold can be specified; contaminants are only detected above a specified
concentration limit (the default limit is zero). Second, a response time can be specified, which accounts for the time needed to verify the presence of contamination (e.g. by field investigation) and then inform the public (the default
response time is zero). The contamination impact is computed at the
time where the response has completed (the detection time plus response
time), which is called the {\itshape effective response time\/}. For
undetected incidents, the effective response time is simply the end of
the contaminant transport simulation. The following illustrates how to
specify these options:
\lstinputlisting[firstline=16,lastline=16]{examples/simple3.sh}% anchor=tso2Impact
This computes impacts for a 60 minute response time, with a 0.1 detection
threshold. Note that the units for {\ttfamily -\/-\/detectionLimit}
are the same as for the MASS values that are specified in the TSG file.

Impacts from multiple ERD files can be combined to generate a single
IMPACT file using the following syntax:
\lstinputlisting[firstline=24,lastline=24]{examples/simple4.sh}% anchor=tso2Impact
Note that the value of 30000000 corresponds to the detection threshold
for the contaminant described in {\ttfamily Net3\_\-1a.erd} and
0.0001 is the detection threshold for the contaminant described in
{\ttfamily Net3\_\-1b.erd}. For example, this can be used to combine
simulation results from different types of contaminants, in which the
ERD files would have been generated from different TSG files. Murray et
al.~\cite{MurBerHar06} use this technique to combine data from different
types of contamination incidents into a single impact metric.

The dvf option specifies that the demands added due to flushing be subtracted out prior to calculating the impact measures. Add this in if the demands creating by simulating flushing are not ``consumed'' - i.e., you don't want the mass included in mass consumed, or other impacts. The time-extent-of-contamination (tec) impact measure integrates the time that each pipe contains contaminated water, rather than just the length of pipe that ever contains contaminated water. The result is in units of ft-hrs.

The species option specifies which species to use to compute impacts.  This option is required for multi-species contamination incidents created by {\ttfamily tevasim}.  For example:
\lstinputlisting[firstline=16,lastline=16]{examples/simple7.sh}% anchor=tso2Impact  

\section{Impact Measures}\label{impacts_impact2a}

After running \doxyref{tevasim}{p.}{tevasimExecutable} command, the
output database, {\ttfamily Net3.erd}, can be used
to compute one or more IMPACT files. An IMPACT
file summarizes the consequence of each contamination incident in a
manner that facilitates optimization. A variety of objective measures
are supported by {\ttfamily tso2Impact} to reflect the different criteria
that decision makers could use in CWS design. For most of these criteria,
there is a {\itshape detected\/} and {\itshape undetected\/} version
of the objective. This difference concerns how undetected contamination
incidents are modeled.

For example, the default time-\/to-\/detection objective, {\bfseries td},
uses the time at which the EPANET simulations are terminated to define
the time for incidents that are not detected. The effect of this is that undetected incidents severely penalize sensor network designs. By contrast, the {\itshape
detected\/} time-\/to-\/detection, {\bfseries dtd}, simply ignores these
incidents (they have impact zero  and do not affect the sensor network design). Sensor placement with the {\itshape
detected\/} objective is somewhat more precise but optimization can be slow.  Ideally, the objective is optimized with a revised formulation that explicitly limits the
fraction of incidents that are not detected by the sensors.  However, in real world applications, the detected metric is typically used without side constraints.

The following objectives are currently supported by {\ttfamily tso2Impact:} 
\begin{itemize}

\item {\bfseries ec} and {\bfseries dec} -\/ The extent of contaminated
in the network. This is the total feet of pipes contaminated by the
effective response time. An entire pipe is considered contaminated if
contaminant enters the pipe at a given time step. For {\bfseries ec},
the extent of contamination of an undetected incident is the extent
of contamination at the point when the simulation terminates, while
undetected contamination incidents are ignored for {\bfseries dec}.

\item {\bfseries mc} and {\bfseries dmc} -\/ The mass of contaminant
consumed by junctions in the network with nonzero demand. For {\bfseries
mc}, the mass of contaminant of an undetected incident is the mass of
contaminant that has left the network via demand at the point when
the simulation terminates, while undetected contamination incidents
are ignored for {\bfseries dmc}. This objective is typically measured
in milligrams (the units used in the TSG file are mg/L). However,
concentrations may also be interpreted; for example, we can treat this
measure as a count of cells for a biological contaminant, where the TSG
measurement is cells/L.

\item {\bfseries nfd} -\/ The number of contamination incidents that are
not detected by any sensor before the contaminant transport simulations
terminate. NOTE: this measure is not affected by the response time option.

\item {\bfseries pe} and {\bfseries dpe} -\/ The number of individuals
exposed to a contaminant. For {\bfseries pe}, the population exposed
for an undetected incident is the population exposed at the point when
the simulation terminates, while undetected contamination incidents are
ignored for {\bfseries dpe}.

\item {\bfseries pd} and {\bfseries dpd} -\/ The number of individuals
that receive a dose of contaminant above a specified threshold. For
{\bfseries pd}, the population dosed by an undetected incident is the
population dosed at the point when the simulation terminates, while for
{\bfseries dpd} the undetected contamination incidents are ignored.

\item {\bfseries pk} and {\bfseries dpk} -\/ The number of individuals
killed by a contaminant. For {\bfseries pk}, the population killed
by an undetected incident is the population killed at the point when
the simulation terminates, while for {\bfseries dpk} the undetected
contamination incidents are ignored.

\item {\bfseries td} and {\bfseries dtd} -\/ The time, in minutes, from
the beginning of a contamination incident until the first sensor detects
it. For {\bfseries td}, the time-\/to-\/detection of an undetected
incident is the time from the start of the incident until the end of
the simulation, while undetected contamination incidents are ignored
for {\bfseries dtd}. NOTE: this measure is not affected by the response
time option.

\item {\bfseries vc} and {\bfseries dvc} -\/ The volume of contaminated
water consumed by junctions in the network with nonzero demand. For
{\bfseries vc}, the volume of contaminated water of an undetected
incident is the volume of contaminated water consumed at the point when
the simulation terminates, while undetected contamination incidents are
ignored for {\bfseries dvc}.

\end{itemize}

These health impact measures are computed with an auxiliary input file, {\bfseries TAI}, that specifies parameters for a health impact model that predicts how a population is affected by exposure to a contaminant. The \doxyref{TAI File}{p.}{formats_taiFile} bio.tai specifies the nature of the contaminant and how it impacts human health. Further, this file specifies the fraction of the volume of water consumed at junctions that is consumed by humans. For example, consider the command line: 
\lstinputlisting[firstline=16,lastline=16]{examples/simple5.sh}% anchor=tso2Impact


\section{Advanced Tools for Large Sensor Placements Problems}\label{impacts_impact3}

In some applications, the size of the IMPACT files is very
large, which can lead to optimization models that cannot
be solved on standard 32-\/bit workstations. SPOT includes
several utilities that are not commonly used to address this
challenge: the \doxyref{scenarioAggr}{p.}{scenarioAggrExecutable}
executable aggregates similar contamination incidents, and the
\doxyref{filter\_\-impacts}{p.}{filter_impactsExecutable} script filters
out contamination incidents that have low impacts.

The \doxyref{scenarioAggr}{p.}{scenarioAggrExecutable} executable reads
an IMPACT file, finds similar incidents, combines them, and writes out
another IMPACT file. This aggregation technique combines two incidents
that impact the same locations in the same order, allowing for the
possibility that one incident continues to impact other locations. For
example, two contamination incidents should travel in the same pattern
if they differ only in the nature of the contaminant, though one may
decay more quickly than the other. Aggregated incidents can be combined
by simply averaging the impacts that they observe and updating the
corresponding incident weight.

For example, consider the command:
\lstinputlisting[firstline=20,lastline=20]{examples/simple6.sh}% anchor=aggr
This creates the files {\ttfamily aggrNet3\_\-mc.impact} and {\ttfamily aggrNet3\_\-mc.impact.prob}; where the {\ttfamily Net3\_\-mc.impact} file has 236 events. The file {\ttfamily aggrNet3\_\-mc.impact} is the new IMPACT file, and the file {\ttfamily aggrNet3\_\-mc.impact.prob} contains the probabilities of the aggregated incidents.

The \doxyref{filter\_\-impacts}{p.}{filter_impactsExecutable} script
reads an impact file, filters out the low-\/impact incidents, rescales
the impact values, and writes out another IMPACT file. The command:
\lstinputlisting[firstline=24,lastline=24]{examples/simple6.sh}% anchor=filter
generates an IMPACT file that contains the incidents whose impacts
(without sensors) are the largest 5\% of the incidents in {\ttfamily
Net3\_\-mc.impact}. Similarly, the {\ttfamily -\/-\/num=k} option selects
the $k$ incidents with the largest impacts, and the option {\ttfamily
-\/-\/threshold=h} selects the incidents with the impacts greater than
or equal to $h$.

The {\ttfamily filter\_\-impacts} command also includes options to rescale
the impact values. The {\ttfamily -\/-\/rescale} option rescales impact
values with a log-\/scale and the {\ttfamily -\/-\/round} option rescales
impact values to rounded log-\/scale values.
